---
title: "Clustering Quality Control - Answer Key"
author:
  - Noor Sohail
  - Will Gammerdinger
date: "2025-07-01"
---

```{r}
#| label: load_libraries
#| echo: false
library(Seurat)
library(tidyverse)
library(ggplot2)
library(cowplot)
load(bzfile("data/seurat_integrated.RData.bz2"))
```

# Exercise 1

Hypothesize the clusters corresponding to each of the different clusters in the table:

| Cell Type | Clusters |
|:---:|:---:|
| CD14+ monocytes | 1, 3 | 
| FCGR3A+ monocytes | 10 |
| Conventional dendritic cells | 14 |
| Plasmacytoid dendritic cells | 16 |
| Marcrophages | - |
| B cells | ? |
| T cells | ? |
| CD4+ T cells | ? |
| CD8+ T cells| ? |
| NK cells | ? |
| Megakaryocytes | ? |
| Erythrocytes | ? |
| Unknown | ? |

## B Cells

```{r}
#| label: B_cells_plot
FeaturePlot(seurat_integrated, 
            reduction = "umap", 
            features = c("CD79A", "MS4A1"), 
            order = TRUE,
            min.cutoff = 'q10', 
            label = TRUE)
```

B cells appear to be clusters 7, 11 and 13.

## T cells

```{r}
#| label: T_cells_plot
FeaturePlot(seurat_integrated, 
            reduction = "umap", 
            features = c("CD3D"), 
            order = TRUE,
            min.cutoff = 'q10', 
            label = TRUE)
```

T cells appear to be clusters 0, 2, 4, 5, 6 and 9. 

## CD4+ T cells

```{r}
#| label: CD4_T_cells_plot
FeaturePlot(seurat_integrated, 
            reduction = "umap", 
            features = c("CD3D", "IL7R", "CCR7"), 
            order = TRUE,
            min.cutoff = 'q10', 
            label = TRUE)
```

CD4+ T cells appear to be clusters 0, 2, 4 and 6. 

## CD8+ T cells

```{r}
#| label: CD8_T_cells_plot
FeaturePlot(seurat_integrated, 
            reduction = "umap", 
            features = c("CD3D", "CD8A"), 
            order = TRUE,
            min.cutoff = 'q10', 
            label = TRUE)
```

CD8+ T cells appear to be clusters 5 and 9. 

## NK cells

```{r}
#| label: NK_cells_plot
FeaturePlot(seurat_integrated, 
            reduction = "umap", 
            features = c("GNLY", "NKG7"), 
            order = TRUE,
            min.cutoff = 'q10', 
            label = TRUE)
```

NK cells appear to be clusters 8 and 12. 

## Megakaryocytes

```{r}
#| label: Megakaryocytes_plot
FeaturePlot(seurat_integrated, 
            reduction = "umap", 
            features = c("PPBP"), 
            order = TRUE,
            min.cutoff = 'q10', 
            label = TRUE)
```

Megakaryocytes appear to be clusters 15. 

## Erythrocytes

```{r}
#| label: Erythrocytes_plot
FeaturePlot(seurat_integrated, 
            reduction = "umap", 
            features = c("HBB", "HBA2"), 
            order = TRUE,
            min.cutoff = 'q10', 
            label = TRUE)
```

Erythrocytes don't appear to be present.

***

| Cell Type | Clusters |
|:---:|:---:|
| CD14+ monocytes | 1, 3 | 
| FCGR3A+ monocytes | 10 |
| Conventional dendritic cells | 14 |
| Plasmacytoid dendritic cells | 16 |
| Marcrophages | - |
| B cells | 7, 11, 13 |
| T cells | 0, 2, 4, 5, 6, 9 |
| CD4+ T cells | 0, 2, 4, 6 |
| CD8+ T cells| 5, 9 |
| NK cells | 8, 12 |
| Megakaryocytes | 15 |
| Erythrocytes | - |

