---
title: "Seurat Cheatsheet - Answer Key"
author:
  - Noor Sohail
  - Will Gammerdinger
date: "2025-12-07"
---

```{r}
#| label: load_data
#| echo: false
library(Seurat)
library(tidyverse)

load(bzfile("data/seurat_integrated.RData.bz2"))
```

# Exercise 1

**What are the last 5 cells barcodes and the last 5 genes in the integrated seurat object.**


```{r}
#| label: print_final_genes_cells
# Barcodes
Cells(seurat_integrated) %>% tail()
# Barcodes
colnames(seurat_integrated) %>% tail()

# Genes
Features(seurat_integrated) %>% tail()
# Genes
rownames(seurat_integrated) %>% tail()
```


# Exercise 2

**What are the last 5 identities for the cells in the integrated seurat object?**

```{r}
#| label: rename_idents
#| echo: false
# Rename all identities
seurat_integrated <- RenameIdents(object = seurat_integrated, 
                                 "1" = "CD14+ monocytes",
                                 "3" = "CD14+ monocytes",
                                 "2" = "Activated T cells")

# These new celltype values are only stored in the idents
# Good practice is to store these changes in a column
seurat_integrated$celltype <- Idents(seurat_integrated)
```

```{r}
#| label: print_final_idents
Idents(seurat_integrated) %>% tail()
```


# Exercise 3

**What are the the 5 least variable genes in the integrated seurat object?**

```{r}
#| label: set_var_features
#| echo: false
# Get list of all variable genes
# Remove variable genes that start with MT-
var_genes <- VariableFeatures(seurat_integrated)
var_genes <- var_genes[!startsWith(var_genes, "MT-")]

# Now we set our vector of gene names back to VariableFeatures()
VariableFeatures(seurat_integrated) <- var_genes
```


```{r}
#| label: show_var_features
VariableFeatures(seurat_integrated) %>% tail()
```

::: callout-note
You may have noticed that these are the same genes from when we were looking at the last 5 genes in our Seurat object. This is due to the fact that we ran integration on the top 3,000 variable genes. As a result, the order of the genes in the integrated dataset follows the same order as the variable features!
:::

# Exercise 4

**What are the dimensions for each assay in the integrated seurat object?**

```{r}
#| label: show_dim_assays
dim(seurat_integrated[["RNA"]])
dim(seurat_integrated[["SCT"]])
dim(seurat_integrated[["integrated"]])
```

::: callout-note
Notice that the number of genes is higher in the RNA assay compared to the integrated object. This goes back to previous note, where only the expression from variable genes is stored in the `integrated` assay.
:::

# Exercise 5

**Show the code to get the entire SCT normalized (data) count matrix.**

```{r}
#| label: get_counts_mtx
#| eval: false
LayerData(seurat_integrated, assay="SCT", layer="data")
```

# Exercise 6

**Show how you would use the `FetchData()` function to generate a dataframe of UMAP_1, UMAP_2, and sample values for each cell.**

```{r}
#| label: fetchdata_exercise
FetchData(seurat_integrated, vars=c("UMAP_1", "UMAP_2", "sample")) %>% head()
```
