---
title: "Marker identification - Answer Key"
author:
  - Noor Sohail
  - Will Gammerdinger
date: "2025-12-07"
---

```{r}
#| label: load_libraries
#| echo: false
library(Seurat)
library(tidyverse)
load(bzfile("data/seurat_integrated.RData.bz2"))
annotations <- read.csv("data/annotation.csv")
```

# Exercise 1

1. In the previous lesson, we identified cluster 10 as FCGR3A+ monocytes by inspecting the expression of known cell markers FCGR3A and MS4A7. Use `FindConservedMarkers()` function to find conserved markers for cluster 10. What do you observe? Do you see FCGR3A and MS4A7 as highly expressed genes in cluster 10?

```{r}
#| label: FindConservedMarkers_cluster_10
# FindConservedMarkers for cluster 10
cluster10_conserved_markers <- FindConservedMarkers(seurat_integrated,
                                                    ident.1 = 10,
                                                    grouping.var = "sample",
                                                    only.pos = TRUE,
                                                    logfc.threshold = 0.25)

# Combine markers with gene descriptions 
cluster10_ann_markers <- cluster10_conserved_markers %>% 
                rownames_to_column(var="gene") %>% 
                left_join(y = unique(annotations[, c("gene_name", "description")]),
                          by = c("gene" = "gene_name"))

# View the top 6 marker genes
head(cluster10_ann_markers)
```

Yes, FCGR3A and MS4A7 are among the top highly expressed genes when performing `FindConservedMarkers()` for cluster 10. This observation confirms cluster 10 as FCGR3A+ monocytes.